### 2022-11-11 submission/manuscript.Rmd
First full draft of Introduction

n_med_sequence_range <- table_1 %>%
  pull(median) %>%
  range() %>% as.integer()

@@

***https://edepot.wur.nl/547087.*** "Log- ratio PCA is designed to give results that are library size- independent. However, as we demonstrated mathematically and with examples based on simulated and real data, log- ratio PCA becomes library size- dependent, if there are many infrequent taxa (many zeroes) and library sizes differ largely. In this situation, the row centring used in log- ratio PCA brings an effect of r (the row mean of the log- transformed counts) in the clr- transformed matrix. Note that this effect is irrespective of whether or not these infrequent taxa are genuine or due to sequencing noise or allocation error. This library size dependence is unexpected in the sense that, after applying the clr, the transformed matrix is free of the effect of the row totals for strictly positive data (yij>0 for all i and j). We additionally demonstrate that library size variability causes a loss in power in dextecting an effect of x with log- ratio RDA. If there is additionally a correlation between treatment and the library size, the type 1 error for detecting the effect of x can be seriously inflated."

***https://www.nature.com/articles/s41522-020-00160-w.*** "An important characteristic of a feature table is that it is typically sparse, sometimes as many as ~90% are zero entries21, which creates a challenge for analyzing rare taxa. A quick and simple strategy to deal with excess zeros is to add a small positive constant (e.g. 1) called pseudo-count14,22 to each cell of the feature table. The addition of a pseudo-count becomes necessary when using methods of analysis that require log transformation of the observed counts. Even though adding a pseudo-count is simple and widely used, the choice of the pseudo-count is ad hoc. Studies have shown that differential abundance or clustering results could be sensitive to the choice of pseudo count23,24. Although different values of pseudo counts have been discussed in the literature23,24,25,26, to the best of our knowledge, there is no consensus on how to choose the optimal value. Other strategies involve modeling zero counts by some probability models21,27."

https://academic.oup.com/bioinformatics/article/34/16/2870/4956011
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6755255/

* Motivation
  * Problem of uneven sampling effort
* What is rarefaction? History, reason for rarefaction
  * Repeated down sampling of datasets to a common number of observations to calculate the average value to ascertain the expected value of a metric for the metric under study; typically richness
  * Control for unneven sampling effort
  * Methods vary in their sensitivity to uneven sampling
  * Compositional data analysis
* Reasons behind "rarefaction is inadmissable"
  * Weird simulation
* Alternative approaches and claims
  * sampling invariance
* Goal of this study

This analysis included 16S rRNA gene sequence data from from 12 studies that characterized the variation in bacterial communities from diverse environments (Table 1). The original studies generated the sequence data by pooling separate PCR products that were generated by amplifying the V4 region of the 16S rRNA gene from the bacterial DNA in multiple samples. Because pooling equimolar quantities of DNA is frought with difficulties, it was common to observe wide variation in the number of sequences in each sample (Table 1 and Figure S1).

@@

Use of relative abundances, normalized counts, variance stabilizing transformations, and centered log ratios have each been recommended as more robust alternatives to rarefaction. Again, the only approach that consistently removed the effects of uneven sampling effort was rarefaction. Most of the recommendations borrow techniques from methods for identifying differential gene expression. Unfortunately, there is an important but underappreciated difference between gene expression and community data. This is the interpretation of unobserved data. For gene expression analysis in a single organism the lack of any sequencing data for a gene would indicate that its expression was below the limit of detection. Sequencing the same organism under multiple conditions would not lead to a seemingly unbouned number of genes in the organims. Rather, the number of genes has a definite limit that is knowable from the genome sequence. With microbiome data, an unobserved sequence could mean that the organism was present, but below the limit of detection or that the organism was missing. Because we have yet to exhaustively sample any community in the same way we have sequenced a single genome it is unreliable to impute the presence of all organisms. Yet, this is exactly what the variance stabilization transformation and most CLR techniques do. This analysis has demonstrated a clear correlation between distances calculated by these methods and sequencing effort. This result is at odds with the claims by others that the distances are scale invariant [@Martino2019]. Again, rarefaction is the most effective and consistent approach to control for uneven sampling effort when calculating beta diversity metrics.

@@

\small

```{r results_f}
table_1 %>%
  arrange(nice_name) %>%
  mutate(
    nice_name = glue("{nice_name}\\ {reference}"),
    total_seqs=format(total_seqs, big.mark=",", trim=TRUE),
    median=format(as.integer(median), big.mark=",", trim=TRUE),
    min=format(min, big.mark=",", trim=TRUE),
    max=format(max, big.mark=",", trim=TRUE),
    range = glue("{min}\\Hyphdash*{max}"),
    n_samples = n_samples) %>%
  select(nice_name, n_samples, total_seqs, median, range, sra_study) %>%
  kableExtra::kable(format="markdown", booktabs=TRUE, escape=F, align="lrrrrr",
    linesep="",
    col.names = linebreak(
      c(
        "\\textbf{Dataset\\nobreakspace{}(Ref)}",
        "\\textbf{Samples}",
        "\\textbf{Total}\n\\textbf{sequences}",
        "\\textbf{Median}\n\\textbf{sequences}",
        "\\textbf{Range of}\n\\textbf{sequences}",
        "\\textbf{SRA study}\n\\textbf{accession}"
      ),
      align="c")
  )
```
\normalsize

>>>>

#n_med_sequence_range <- table_1 %>%
#  pull(median) %>%
#  range() %>% as.integer()

@@

The ability to generate millions sequence reads from 16S rRNA gene amplicons and metagenomic datasets has allowed researchers to multiplex multiple samples on the same sequencing run by pooling separate PCRs that can be deconvoluted later based on index (aka barcode) sequences that embeded into the primer sequence [@Kozich2013; @Caporaso2010]. Unfortunately, it is common to observe variation in the number of sequence reads from each sample vary by as much as 100-fold (e.g, see Figure S1). This occurs because pooling of DNA from multiple PCRs is frought with numerous opportunities for technichal errors to compound leading to a skewed distribution. Aside from developing better methods of pooling DNA, the question of how to control for uneven sampling effort in microbial ecology studies has become controversial.

The practice of rarefaction has been commonly used in ecology for more than 50 years as a tool to control for uneven sampling effort across experimental replicates [@Sanders1968; @Hurlbert1971]. Microbial ecologists have used it to compare 16S rRNA gene sequence data for the past 25 years [@Dunbar1999; @Hughes2001; @Moyer1998]. When data are rarefied, the investigator selects a desired threshold of sampling effort and removes any samples below that threshold. They then randomly select that many sequences, with replacement from each sample. Based on the obseved sequence counts, the researcher can then calcuate alpha diversity metrics including richness and diversity indices or beta divesity metrics such as a Jaccard or Bray-Curtis dissimilarity index. I refer to this single sampling as a subsample; this method is implemented as the sub.sample function in mothur [@Schloss2009] and the rrarefy function in the vegan R package [@Oksanen2022]. Rarefaction repeats the subsampling a large number of times (e.g., 100 or 1,000) and calculates the mean of the alpha or beta diersity metric over those subsamplings; rarefaction is implemented in mothur using the summary.single and dist.shared functions [@Schloss2009] and with the vegan R package using the rarefy or avgdist fuctions [@Oksanen2022]. Rarefaction effectively tells a researcher what an alpha or beta diversity metric would have been for a collection of samples if they were all evenly sampled. Although a closed form equation exists to calcualte the expected richness [@Hurlbert1971], it is computationally easier to empirically calculate rarefied richness and other alpha and beta diversity metrics.

In 2014, McMurdie and Holmes [@McMurdie2014] announced that rarefaction of microbial community data is "statistically inadmissible" because it omits valid data. In their simulations, they observed that the use of rarefaction reduced the statistical power to compare communities based on beta diversity metrics. The communities used in this study were simulated by mixing operational taxonomic unit (OTU) frequencies from marine and human fecal communities, which were then sampled to generate 40 replicate communities each sampled to different depths. The simulation with the greatest depth of coverage obtained an average of 10,000 sequences per sample. Analysis of the R code that they used indicates that the simulations were actually only based on a single subsample and not truly rarefied data. Furthermore, the authors removed any OTU from the simulated communities that was not observed in three or more samples. This is a frequently used approach for analyzing microbiome data [@Bokulich2013]. Unfortunately, it removes valid data, skews the distribution of the community, and removes true patchiness from the representation of the communities [@Schloss2020]. Finally, the analysis did not include analysis of alpha diersity metrics and chose to assess beta diversity metrics based on clustering quality of samples rather than using the more frequently used approach of employing non-parametric multivariate analysis of variance [@McArdle2001]. Regardless of these caveats, McMurdie and Holmes recommended the use of variance stabilizing transformations such as those applied for gene expression analysis such as that found in the DeSeq R package.

Beyond the critiques from McMurdie and Holmes others have developed alternative approaches to controlling for uneven sampling effort in amplicon sequencing studies. For alpha diversity metrics, Willis used toy datasets to demonstrate that one could estimate the richness for each sample in a dataset and use those values for statistical comparisons [@Willis2019]. Non-parametric estimators of richness and diversity [@Chao2003; @Chao2016] and parametric estimators of richness [Willis2015] that have been used in microbial ecology studies. For beta diversity metrics at least four approaches have been pursued. First, one could use relative abundance values where the observed number of sequences in an OTU is divided by the total number of sequences in the sample [@Lin2020]. Second, normalization strategies have been developed where the relative abundance is mutiplied by the size of the minimum desired sampling effort and fractional values are reapportioned to the OTUs to obtain integer values [@Beule2020; @Paulson2013]. Third, a variety of center log-ratio methods have been developed where the compositional nature of the OTU counts is removed and used to calculated Euclidean distances (aka Aitchison distances) [@Quinn2019; @Paulson2013; @Lin2020; @Martino2019; @Costea2014. This strategy is purported to control for uneven sampling effort [@Martino2019; @Quinn2018]; however, some have noticed that this feature breaks down under certain conditions [@teBeest2021]. Finally, as mentioned above, variance stabilization transformations have been recommended to generate values that can be used to calcualte Euclidean distances [@McMurdie2014].

The ongoing controversy over the use of rarefaction and the development of numerous strategies to control for uneven sampling effort caused me to question how these methods compared to each other. My analysis included 16S rRNA gene sequence data from from 12 studies that characterized the variation in bacterial communities from diverse environments (Table 1 and Figure S1). The sequences were assigned to OTUs using a standard pipeline and their frequencies and the number of sequences found in each sample were used to generate simulated communities and treatment effects. For each dataset and simulation, 100 replicate datasets were generated and used as input to each of the strategies for controlling for uneven sampling effort. My overall conclusion was that rarefaction outperformed the alternative strategies.

@@

Use of relative abundances, normalized counts, variance stabilizing transformations, and centered log ratios have each been recommended as more robust alternatives to rarefaction. Again, the only approach that consistently removed the effects of uneven sampling effort was rarefaction. Most of the recommendations borrow techniques from methods for identifying differential gene expression. Unfortunately, there is an important but underappreciated difference between gene expression and community data. This is the interpretation of unobserved data. For gene expression analysis in a single organism the lack of any sequencing data for a gene would indicate that its expression was below the limit of detection. Sequencing the same organism under multiple conditions would not lead to a seemingly unbouned number of genes in the organims. Rather, the number of genes has a definite limit that is knowable from the genome sequence. With microbiome data, an unobserved sequence could mean that the organism was present, but below the limit of detection or that the organism was missing. Because we have yet to exhaustively sample any community in the same way we have sequenced a single genome it is unreliable to impute the presence of all organisms. Yet, this is exactly what the variance stabilization transformation and most CLR techniques do. This analysis has demonstrated a clear correlation between distances calculated by these methods and sequencing effort. This result is at odds with the claims by others that the distances are scale invariant [@Martino2019; @Quinn2018]. Again, rarefaction is the most effective and consistent approach to control for uneven sampling effort when calculating beta diversity metrics.

@@

\footnotesize

```{r results_f}
table_1 %>%
  arrange(nice_name) %>%
  mutate(
    nice_name = glue("{nice_name}\\ {reference}"),
    total_seqs=format(total_seqs, big.mark=",", trim=TRUE),
    median=format(as.integer(median), big.mark=",", trim=TRUE),
    mean=format(as.integer(mean), big.mark=",", trim=TRUE),
    min=format(min, big.mark=",", trim=TRUE),
    max=format(max, big.mark=",", trim=TRUE),
    range = glue("{min}\\Hyphdash*{max}"),
    n_samples = n_samples) %>%
  select(nice_name, n_samples, total_seqs, median, mean, range, sra_study) %>%
  kbl(format="markdown", booktabs=TRUE, escape=F, align="lrrrrrr",
    linesep="",
    col.names = linebreak(
      c(
        "\\textbf{Dataset\\nobreakspace{}(Ref)}",
        "\\textbf{Samples}",
        "\\textbf{Total}\n\\textbf{sequences}",
        "\\textbf{Median}\n\\textbf{sample size}",
        "\\textbf{Mean}\n\\textbf{sample size}",
        "\\textbf{Range of}\n\\textbf{sample sizes}",
        "\\textbf{SRA study}\n\\textbf{accession}"
      ),
      align="c")
  )
```
\normalsize
